A comparison of design-based and model-based approaches for finite population spatial sampling and inference.

### 2022-05-19 inst/manuscript/manuscript.Rmd
minor updates

The rest of the manuscript is organized as follows. In Section \ref{sec:background}, we introduce and provide relevant background for design-based and model-based approaches to finite population spatial sampling and inference. In Section \ref{sec:mm}, we describe how we intend to compare performance of the approaches using simulated and real data. The real data is from the United States Environmental Protection Agency's 2012 National Lakes Assessment (NLA) [@USEPA2012NLA]. In Section \ref{sec:results}, we present analysis results for the simulated data and NLA data. And in Section \ref{sec:discussion}, we end with a discussion and provide directions for future research.

@@

We previously mentioned that the design-based approach can be used to select spatially balanced samples. Spatially balanced samples are useful because parameter estimates from these samples tend to vary less than parameter estimates from samples lacking spatial balance [@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @wang2013design; @benedetti2017spatiallyreview]. The first spatially balanced sampling algorithm to see widespread use was the generalized random tessellation stratified (GRTS) algorithm [@stevens2004spatially]. After the GRTS algorithm was developed, several other spatially balanced sampling algorithms emerged, including stratified sampling with compact geographical strata [@walvoort2010r], the local pivotal method [@grafstrom2012spatially; @grafstrom2018spatially], spatially correlated Poisson sampling [@grafstrom2012spatiallypoisson], balanced acceptance sampling [@robertson2013bas], within-sample-distance sampling [@benedetti2017spatially], and Halton iterative partitioning sampling [@robertson2018halton]. In this manuscript, we select spatially balanced samples using the GRTS algorithm because it is readily available in the `spsurvey` \textbf{\textsf{R}} package [@dumelle2022spsurvey] and naturally accommodates finite and infinite sampling frames, unequal inclusion probabilities, and replacement units. Replacement units are additional population units that can be sampled when a population unit originally selected can no longer be sampled. A couple of reasons why an originally selected site can no longer be sampled include its location being physically inaccessible or it is on private land that the researcher does not have permission to access.

@@

In each of the 36 simulation scenarios, there were 2000 independent simulation trials. Within each trial, a population was simulated according to the specifications of the particular simulation scenario (for the random location layout, locations were simulated separately for each trial). Next, SRS and GRTS samples were selected. Then, design-based and model-based inferences were used to estimate (design-based) or predict (model-based) the realized mean and construct 95% confidence (design-based) or 95% prediction (model-based) intervals. With model-based inference, covariance parameters and the realized mean were estimated (using REML) separately for each trial.  After all 2000 trials, we summarized the long-run performance of the sampling-inference combination in each scenario by calculating mean bias, root-mean-squared error, and interval coverage. Mean bias was taken as the average deviation between each trial's estimated (or predicted) mean ($\hat{\mu}_i$) and its realized mean ($\mu_i$): $\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu_i)$, where $i$ indexes the simulation trials. Because each trial had different realized populations, $\mu_i$ changed with $i$. Root-mean-squared error was taken as the square root of the average squared deviation between each trial's estimated (or predicted) mean and its realized mean: $\sqrt{\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu_i)^2}$. Interval coverage was taken as the proportion of simulation trials where the realized mean was contained in its 95% confidence (or prediction) interval. These intervals were constructed using the normal distribution -- justification comes from the asymptotic normality of means via the central limit theorem (under some assumptions). Quantifying these metrics is important because together, they give us an idea of the accuracy (mean bias), spread (RMSE), and validity (interval coverage) of the sampling-inference combinations.

>>>>

The rest of the manuscript is organized as follows. In Section \ref{sec:background}, we introduce and provide relevant background for design-based and model-based approaches to finite population spatial sampling and inference. In Section \ref{sec:mm}, we describe how we intend to compare performance of the approaches using simulated and real data. The real data is from the United States Environmental Protection Agency's 2012 National Lakes Assessment (NLA) [@USEPA2012NLA]. In Section \ref{sec:results}, we present analysis results for the simulated data and real data. And in Section \ref{sec:discussion}, we end with a discussion and provide directions for future research.

@@

We previously mentioned that the design-based approach can be used to select spatially balanced samples. Spatially balanced samples are useful because parameter estimates from these samples tend to vary less (be more precise) than parameter estimates from samples lacking spatial balance [@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @wang2013design; @benedetti2017spatiallyreview]. The first spatially balanced sampling algorithm to see widespread use was the generalized random tessellation stratified (GRTS) algorithm [@stevens2004spatially]. After the GRTS algorithm was developed, several other spatially balanced sampling algorithms emerged, including stratified sampling with compact geographical strata [@walvoort2010r], the local pivotal method [@grafstrom2012spatially; @grafstrom2018spatially], spatially correlated Poisson sampling [@grafstrom2012spatiallypoisson], balanced acceptance sampling [@robertson2013bas], within-sample-distance sampling [@benedetti2017spatially], and Halton iterative partitioning sampling [@robertson2018halton]. In this manuscript, we select spatially balanced samples using the GRTS algorithm because it is readily available in the `spsurvey` \textbf{\textsf{R}} package [@dumelle2022spsurvey] and naturally accommodates finite and infinite sampling frames, unequal inclusion probabilities, and replacement units. Replacement units are additional population units that can be sampled when a population unit originally selected can no longer be sampled. A couple of reasons why an originally selected site can no longer be sampled include its location being physically inaccessible or it is on private land that the researcher does not have permission to access.

@@

In each of the 36 simulation scenarios, there were 2000 independent simulation trials. Within each trial, a population was simulated according to the specifications of the particular simulation scenario (for the random location layout, locations were simulated separately for each trial). Next, a random SRS sample and a random GRTS sample were selected. Then, design-based and model-based inferences were used to estimate (design-based) or predict (model-based) the realized mean and construct 95% confidence (design-based) or 95% prediction (model-based) intervals. With model-based inference, covariance parameters and the realized mean were estimated (using REML) separately for each trial.  After all 2000 trials, we summarized the long-run performance of the sampling-inference combination in each scenario by calculating mean bias, root-mean-squared error, and interval coverage. Mean bias was taken as the average deviation between each trial's estimated (or predicted) mean ($\hat{\mu}_i$) and its realized mean ($\mu_i$): $\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu_i)$, where $i$ indexes the simulation trials. Because each trial had a different realized population, $\mu_i$ changed with $i$. Root-mean-squared error was taken as the square root of the average squared deviation between each trial's estimated (or predicted) mean and its realized mean: $\sqrt{\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu_i)^2}$. Interval coverage was taken as the proportion of simulation trials where the realized mean was contained in its 95% confidence (or prediction) interval. These intervals were constructed using the normal distribution -- justification comes from the asymptotic normality of means via the central limit theorem (under some assumptions). Quantifying these metrics is important because together, they give us an idea of the accuracy (mean bias), spread (RMSE), and validity (interval coverage) of the sampling-inference combinations.

### 2022-05-19 inst/manuscript/manuscript.tex
minor updates

The rest of the manuscript is organized as follows. In Section
\ref{sec:background}, we introduce and provide relevant background for
design-based and model-based approaches to finite population spatial
sampling and inference. In Section \ref{sec:mm}, we describe how we
intend to compare performance of the approaches using simulated and real
data. The real data is from the United States Environmental Protection
Agency's 2012 National Lakes Assessment (NLA) (USEPA, 2012). In Section
\ref{sec:results}, we present analysis results for the simulated data
and NLA data. And in Section \ref{sec:discussion}, we end with a
discussion and provide directions for future research.

@@

We previously mentioned that the design-based approach can be used to
select spatially balanced samples. Spatially balanced samples are useful
because parameter estimates from these samples tend to vary less than
parameter estimates from samples lacking spatial balance (Barabesi and
Franceschi, 2011; Benedetti et al., 2017; Grafström and Lundström, 2013;
Robertson et al., 2013; Stevens and Olsen, 2004; Wang et al., 2013). The
first spatially balanced sampling algorithm to see widespread use was
the generalized random tessellation stratified (GRTS) algorithm (Stevens
and Olsen, 2004). After the GRTS algorithm was developed, several other
spatially balanced sampling algorithms emerged, including stratified
sampling with compact geographical strata (Walvoort et al., 2010), the
local pivotal method (Grafström et al., 2012; Grafström and Matei,
2018), spatially correlated Poisson sampling (Grafström, 2012), balanced
acceptance sampling (Robertson et al., 2013), within-sample-distance
sampling (Benedetti and Piersimoni, 2017), and Halton iterative
partitioning sampling (Robertson et al., 2018). In this manuscript, we
select spatially balanced samples using the GRTS algorithm because it is
readily available in the \texttt{spsurvey} \textbf{\textsf{R}} package
(Dumelle et al., 2022) and naturally accommodates finite and infinite
sampling frames, unequal inclusion probabilities, and replacement units.
Replacement units are additional population units that can be sampled
when a population unit originally selected can no longer be sampled. A
couple of reasons why an originally selected site can no longer be
sampled include its location being physically inaccessible or it is on
private land that the researcher does not have permission to access.

@@

In each of the 36 simulation scenarios, there were 2000 independent
simulation trials. Within each trial, a population was simulated
according to the specifications of the particular simulation scenario
(for the random location layout, locations were simulated separately for
each trial). Next, SRS and GRTS samples were selected. Then,
design-based and model-based inferences were used to estimate
(design-based) or predict (model-based) the realized mean and construct
95\% confidence (design-based) or 95\% prediction (model-based)
intervals. With model-based inference, covariance parameters and the
realized mean were estimated (using REML) separately for each trial.
After all 2000 trials, we summarized the long-run performance of the
sampling-inference combination in each scenario by calculating mean
bias, root-mean-squared error, and interval coverage. Mean bias was
taken as the average deviation between each trial's estimated (or
predicted) mean (\(\hat{\mu}_i\)) and its realized mean (\(\mu_i\)):
\(\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu_i)\), where \(i\)
indexes the simulation trials. Because each trial had different realized
populations, \(\mu_i\) changed with \(i\). Root-mean-squared error was
taken as the square root of the average squared deviation between each
trial's estimated (or predicted) mean and its realized mean:
\(\sqrt{\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu_i)^2}\).
Interval coverage was taken as the proportion of simulation trials where
the realized mean was contained in its 95\% confidence (or prediction)
interval. These intervals were constructed using the normal distribution
-- justification comes from the asymptotic normality of means via the
central limit theorem (under some assumptions). Quantifying these
metrics is important because together, they give us an idea of the
accuracy (mean bias), spread (RMSE), and validity (interval coverage) of
the sampling-inference combinations.

>>>>

The rest of the manuscript is organized as follows. In Section
\ref{sec:background}, we introduce and provide relevant background for
design-based and model-based approaches to finite population spatial
sampling and inference. In Section \ref{sec:mm}, we describe how we
intend to compare performance of the approaches using simulated and real
data. The real data is from the United States Environmental Protection
Agency's 2012 National Lakes Assessment (NLA) (USEPA, 2012). In Section
\ref{sec:results}, we present analysis results for the simulated data
and real data. And in Section \ref{sec:discussion}, we end with a
discussion and provide directions for future research.

@@

We previously mentioned that the design-based approach can be used to
select spatially balanced samples. Spatially balanced samples are useful
because parameter estimates from these samples tend to vary less (be
more precise) than parameter estimates from samples lacking spatial
balance (Barabesi and Franceschi, 2011; Benedetti et al., 2017;
Grafström and Lundström, 2013; Robertson et al., 2013; Stevens and
Olsen, 2004; Wang et al., 2013). The first spatially balanced sampling
algorithm to see widespread use was the generalized random tessellation
stratified (GRTS) algorithm (Stevens and Olsen, 2004). After the GRTS
algorithm was developed, several other spatially balanced sampling
algorithms emerged, including stratified sampling with compact
geographical strata (Walvoort et al., 2010), the local pivotal method
(Grafström et al., 2012; Grafström and Matei, 2018), spatially
correlated Poisson sampling (Grafström, 2012), balanced acceptance
sampling (Robertson et al., 2013), within-sample-distance sampling
(Benedetti and Piersimoni, 2017), and Halton iterative partitioning
sampling (Robertson et al., 2018). In this manuscript, we select
spatially balanced samples using the GRTS algorithm because it is
readily available in the \texttt{spsurvey} \textbf{\textsf{R}} package
(Dumelle et al., 2022) and naturally accommodates finite and infinite
sampling frames, unequal inclusion probabilities, and replacement units.
Replacement units are additional population units that can be sampled
when a population unit originally selected can no longer be sampled. A
couple of reasons why an originally selected site can no longer be
sampled include its location being physically inaccessible or it is on
private land that the researcher does not have permission to access.

@@

In each of the 36 simulation scenarios, there were 2000 independent
simulation trials. Within each trial, a population was simulated
according to the specifications of the particular simulation scenario
(for the random location layout, locations were simulated separately for
each trial). Next, a random SRS sample and a random GRTS sample were
selected. Then, design-based and model-based inferences were used to
estimate (design-based) or predict (model-based) the realized mean and
construct 95\% confidence (design-based) or 95\% prediction
(model-based) intervals. With model-based inference, covariance
parameters and the realized mean were estimated (using REML) separately
for each trial. After all 2000 trials, we summarized the long-run
performance of the sampling-inference combination in each scenario by
calculating mean bias, root-mean-squared error, and interval coverage.
Mean bias was taken as the average deviation between each trial's
estimated (or predicted) mean (\(\hat{\mu}_i\)) and its realized mean
(\(\mu_i\)): \(\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu_i)\),
where \(i\) indexes the simulation trials. Because each trial had a
different realized population, \(\mu_i\) changed with \(i\).
Root-mean-squared error was taken as the square root of the average
squared deviation between each trial's estimated (or predicted) mean and
its realized mean:
\(\sqrt{\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu_i)^2}\).
Interval coverage was taken as the proportion of simulation trials where
the realized mean was contained in its 95\% confidence (or prediction)
interval. These intervals were constructed using the normal distribution
-- justification comes from the asymptotic normality of means via the
central limit theorem (under some assumptions). Quantifying these
metrics is important because together, they give us an idea of the
accuracy (mean bias), spread (RMSE), and validity (interval coverage) of
the sampling-inference combinations.
